---
title: "DSC 202 Assignment 3"
author: "Jiordani Etienne"
date: "2024-03-06"
output:
  pdf_document: default
  html_document: default
  word_document: default
---
This assignment is about visualizing a dataset of my choosing with great detail, paying attention to the 5 Principles of Accessibility. The principles include:
1. Add Alt Text

2. Takeaway Title

3. Label Data

4. Color Contrast

5. Use White Space

I made a time series model of the proportion of people with a bachelor's degree or higher by gender.

I used the dataset named College_Time.

```{r eval = TRUE, echo = TRUE}
#read in packages
library(tidyverse)
library(ggplot2)
library(readxl)
library(rmarkdown)
```

```{r eval = TRUE, echo = TRUE}
#read in the dataset
col_att <- read_excel('College_Time.xlsx')
```
The dataset contains time series data of the proportion of people who have a bachelor's degree or higher who are 25 years old and older. The dataset goes from 2013 until 2021, and it has two variables, one for men, and the other for women.

The dataset was pulled from Sage Data, which is a database; Sage Data got the data from the World Bank.

```{r eval = TRUE, echo = TRUE}
#look at paged table
paged_table(col_att)
```


```{r eval = TRUE, echo = TRUE}
#assign colors to list
colors <- c("Women" = "#EF0107", "Men" = "#0437F2")
```

```{r eval = TRUE, echo = TRUE}
#make line chart
col_tim <- ggplot(col_att) +
  geom_line(aes(x = Time, y = Female_College, group = 1,  color = "Women"), linewidth = 2) +
  geom_line(aes(x = Time, y = Male_College, group = 1, color = "Men"), linetype = 6, linewidth = 2) +
  labs(title = "Line Chart of the Percentage of People with at Least a Bachelor's", 
       x = "Time (in years)", y = "% of 25+ Year Olds with at Least a Bachelor's,", color = "Legend") +
  scale_color_manual(values = colors) + #adds legend from object named "colors"
  annotate("text", x = 4, y = 34.5, label = "Women") + #makes label for Women
  annotate("text", x = 4, y = 34, label = "|", size = 5) + 
  annotate("text", x = 6, y = 34, label = "|", size = 5) + 
  annotate("text", x = 6, y = 33.5, label = "Men",) #makes label for Men
plot(col_tim)
```

The line chart shows that since 2014, the proportion of women who have a bachelors degree or higher is more than men. In addition, Since 2020, there is a decline in college attendees of on average 2.5 percentage points, which maybe due to the Covid-19 lockdown. In 2021, men with at least a bachelor's degree are sitting at around 33.9% and women at 36.1%.