---
title: "DSC 406 Week 11"
author: "Jiordani Etienne"
date: "2024-04-17"
output: html_document
---
I used the psych package with the NHL file. I renamed a variable and remove missing information. With the psych package, I made a signifcance and correlation matrix of all the variables. I also made a heatmap of the variables and dendrograms with machine learning, one with base R and the other with the pheatmap package. I made a density plot matrix to show how the data is distributed between points and wins. I made a dendrogram with four clusters and colored them. The dendrogram used average distance and it's looking at the relationship between the hockey teams.

Reading in the packages
```{r eval = TRUE, echo = TRUE}
library(tidyverse)
library(rmarkdown)
library(psych)
```

Below is where I read in the dataset
```{r eval = TRUE, echo = TRUE}
#read in the dataset
ds <- read.csv("NHL.csv")
head(ds)
```

rename column name X to Team. 
```{r eval = TRUE, echo = TRUE}
#renames column name to Team
newdata = 
  ds %>% na.omit() %>% rename(Team = X)
```

Checking out the dataset
```{r eval = TRUE, echo = TRUE}
#looking at the dataset
glimpse(newdata)
```

Makes a correlation matrix and significance matrix between all of the variables.
```{r eval = TRUE, echo = TRUE}
#checks if p-values are significant, uses .05 as alpha
newdata[,3:10] %>% round(2) %>% corr.test(alpha = 0.05)
```

Below is a pearson correlation between all fo the variables
```{r eval = TRUE, echo = TRUE}
#makes pearson correlation between all variables
correlation <- cor(newdata[,3:10])
```

Makes a heatmapt of correlation matrix. We needed to calculated the correlation matrix first.The light colors has weak coorrelation, dark colors indicate a strong correlation.
```{r eval = TRUE, echo = TRUE}
#makes a correlatin matrix from the correlation object
library(pheatmap)
pheatmap(correlation, display_numbers = TRUE) #display_numbers shows numbers. Just delete display_numbers to get rid of numbers
```

Creates a heatmap with Base R
```{r eval = TRUE, echo = TRUE}
#makes a heatmap
heatmap(correlation)
```

Makes a correlation matrix and density plot with the GGally package. OL looks like it's the most normally distributed, and PTS and GF seem to have a positive relationship.
```{r eval = TRUE, echo = TRUE}
#makes a correlation and density plot matrix
library(GGally)
newdata[,5:7] %>% ggpairs()
```

Makes a dendrogram of all the teams using the average linkage method
```{r eval = TRUE, echo = TRUE}
library(dendextend)

distances <- dist(newdata) #calculates the distance
dendro <- hclust(distances, method = "average") #uses average linkage method and name dendrogram: dendro
#can call avaerage ave
plot(dendro, hang = -1, labels = newdata$Team) #hang organizes dendrogram
```

Makes a dendrogram for teams separated by 4 colored branches using the average linkage method.
```{r eval = TRUE, echo = TRUE}
graph <- as.dendrogram(hclust(distances, method = "ave"))

labels(graph) <- paste(newdata$Team)[order.dendrogram(graph)] #organizes labels

dend <- color_branches(graph, k = 4, col=c("#CD0000", "#548B54", "#00008B", "#FF00FF"), #adds colors

groupLabels = TRUE) #shows labels

par(cex = 0.7)
plot(dend)
```

