---
title: "Incidents Spatial"
author: "Jiordani Etienne"
date: "2024-02-14"
output: html_document
---
In this week, we I learned how to make a map, subset data, make new variable, and make a histogram. I also showed all the completed rows from incomplete data.

We are reading in the dataset
```{r}
mydata <- read.csv("incidents.csv")
```

```{r}
library(rmarkdown)
library(leaflet)
library(tidyverse)
```


Makes a new variable by renaming them as 1 or 2, and if neither 3 because you don't feel like categorizing all of the other categories.
```{r}
#make a new variable by renaming them as 1 or 2, if neither, then 3
#.default gets rid of all of the other variables that you don't want to categories
mydata %>%
  mutate(new_variable = recode(district,
                               "North" = 1,
                               "Downtown" = 2,
                               .default = 3)) %>%
  select(district, new_variable)
paged_table(mydata)
```

make new variables with longitude and latitude byt narrowing it down to Downtown to see results grouped by years
```{r}
#make new variables with longitude and latitude but only looking at Downtown, seeing results grouped by years
mydata %>%
  filter(district == 'Downtown') %>%
  group_by(reported_year) %>%
  mutate(mean_lat = mean(latitude)) %>%
  mutate(mean_lon = mean(longitude))
```

```{r}
mydata <- read.csv("incidents.csv")
```

We're splitting the dataset based on the reported year
```{r}
#splits big dataset by reported year
newdata <- split(mydata, f = mydata$reported_year)
```

Since the data is split by reported year, we're calling the dataset with the year 2017.
```{r}
#use double brackets to use the piece of the dataset you split up
newdata[['2017']]
```

Now we're naming the dataset of 2017
```{r}
a <- newdata[['2017']]
```

We're getting the relevant variables of dataset a
```{r}
#anaother way to select columns
a <- a[,12:19]
```

We're making a histogram of the reported hours with 10 bars
```{r}
#make a histogram of reported hour
hist(a$reported_hour, col = heat.colors(10))
```

We're replacing 0 values with NA and seeing all of the completed rows
```{r}
#enter missing values as 0
a[a == 0.00000] <- NA

#show completed rows
rowscomplet <- complete.cases(a)
a = a[rowscomplet,]

paged_table(a)
```

We're selecting the variables with only the coordinates.
```{r}
#complete code before the last chunk
mypoints <- a[,7:8]
head(mypoints)
```

We're mapping with the leaflet package. So we get the map from Google Maps, we set the view and plot the points on the maps
```{r}
#read in leaflet
#go back to notes to complete code
library(leaflet)

mypoints %>% leaflet() %>% addTiles() %>% #gets map from Google Maps
  setView(lng = -78.7, lat = 35.7, zoom = 10) %>%
  addCircleMarkers(data = a, radius = 1, color = "red")
```

