---
title: "EITC Project"
author: "Jiordani Etienne"
date: "12/27/23"
output:
  word_document: default
  html_document: default
  pdf_document: default
---
# Effectiveness of the EITC Program on Low-Income Workers and Single Mothers

Research Question: Is the EITC program effective for single mothers, low-income workers, and uneducated workers?

## Introduction
The EITC is a tax credit for low to middle class workers. Eligibility for the EITC depends on family size, income, age, employment, and more. The tax credit is meant to increase wages while reducing taxes paid by low-income workers (Philips 2001). The additional money EITC recipients get can be used for necessities of life like groceries, paying bills, education, children, etc. 

The process of getting the EITC is done by filling out a tax return form, which might be stressful and difficult to complete, especially with low-income workers (Linos et al. 2022). The amount individuals receive in their tax returns depends on how much they make; the tax return increases at first, then it starts to decrease at each addition amount of the filer's salary (Philips 2001 and Linos et all 2022).

Poor people who are in dire need for financial support lack in resources to get support. Research has found that very poor parents, that are on the bottom half of what is considered the federal poverty level, have no or very little knowledge of the EITC program Philips 2001). Furthermore, Philips (2001) found that poor unmarried parents were more likely to have known heard of and received assistance from the EITC program than poor married parents.

Single mothers are by far one of the most important demographics that receive the tax credit. In fact, single mothers have received two-thirds of the the tax credit from 1984 to 1996 (Meyer and Rosenbaum 1999). The tax credit is important to single mothers because of social norms and policies. Most single mothers are single due to a divorce, at which the court system favors women and will most often put the children under the mother's custody. Due to social norms, there are more women who do not possess a post-secondary education than men, which implies that more women choose to be housewives. As a result, many single mothers have to manage working at a job while taking care of her family at the same time.

Research shows that single mothers and single childless women have similar wages (Meyer and Rosenbaum 1999). Meyer and Rosenbaum (1999) also found that single mothers are often older and less educated than childless single women. Therefore, single mothers might be less productive and live more destructive lifestyles compared to childless single women. However, when taxes decrease, and welfare benefits decrease, the employment rate of single women increases (Meyer and Rosenbaum 1999). Moreover, the employment rate increases for single women when there is an increase in welfare benefits if they work, which is due to lower implicit income taxes.

A study found Single women who are high school dropouts tend have a greater increase in employment when a reduction of taxes occur compared to single women with a high school education or higher (Meyer and Rosenbaum 1999).

The EITC is a powerful tool to alleviate financial burdens of paying for bills, education, and children of low-income workers, and promote productivity  and stability for single mothers.

The data is dated to be in 2022, and was pulled from the Bureau of Economic Analysis, U.S Department of Agriculture, and the Census Bureau. The data was compiled in an Excel spreadsheet, then read in to R.

The purpose of the code chunk below is to create a PDF file of the R markdown file.
```{r eval = TRUE, echo = TRUE}
#installs knitr package
library(knitr)

#installs tinytex
library(tinytex)
```


In the following code chunk the data is being read in R.
```{r eval = TRUE, echo = TRUE}
#installs readxl package
library(readxl)

#installs tidyverse package
library(tidyverse)

#reads in EITC_Project Excel file
EITC_Proj <- read_excel('EITC_Project.xlsx')
```

All of the variables are tested using a multiple linear regression model below. The response variable is the EITC variable. The explanatory variables are Income>50k, Single_Women, HS_25, Bachelors_25, Female_Head, Research_Park, and EmpRt.

### What does each of the variables mean?

**State** -- A categorical variable that matches with the other numerical corresponding variables.

**EITC** -- A numerical response variable that indicates how much a state spent on EITC.

**Income50k** -- A numerical explanatory variable that indicates the proportion of people who make less than $50,000 or less in each state.

**Single_Women** -- A numerical explanatory variable that indicates the proportion of women 15 years old and older who are single by either being widowed, divorced, or never married in each state.

**HS_25** -- A numerical explanatory variable of the proportion of individuals with at least a high school education or more in each state.

**Bachelors_25** -- A numerical explanatory variable of the proportion of individuals with a bachelors degree or higher in each state.

**Female_Head** -- A numerical explanatory variable of the proportion of all families that have an unmarried female as the head of a household.

**Research_Park** -- A dummy explanatory variable of the whether or not a state has a research park. 1 indicates the state has a research park, and 0 indicates the state does not have a research park.

**Median_Income** -- A numerical explanatory variable of the median income of each state.

**Poverty_LFPR** -- A numerical explanatory variable of the labor force participation rate of the poor of each state.

**SNAP_Part** -- A numerical explanatory variable of the number of participants in the Supplemental Nutrition Assistance Program of each state.

**LFPR** -- A numerical explanatory variable of the percent of people participating in the labor force in each state.

Below is where the summary statistics is made to test if the explanatory variables correlate with the amount of refunded tax the EITC has in each state.
```{r eval = TRUE, echo = TRUE}
#makes the multiple linear regression model
multi_lin <- lm(EITC ~ Income50k + Single_Women + HS_25 +Bachelors_25 + Female_Head + Research_Park + Median_Income + Poverty_LFPR + SNAP_Part + LFPR, data = EITC_Proj)
print(multi_lin)

#makes ANOVA table
stat_sum <- anova(multi_lin)
print(stat_sum)

#gets correlation coefficient of EITC and Single_Women
corl_Wo <- cor(EITC_Proj$EITC, EITC_Proj$Single_Women)
print(corl_Wo)

#gets correlation coefficient of EITC and HS_25
corl_HS <- cor(EITC_Proj$EITC, EITC_Proj$HS_25)
print(corl_HS)

#gets correlation coefficient of EITC and Bachelors_25
corl_Ba <- cor(EITC_Proj$EITC, EITC_Proj$Bachelors_25)
print(corl_Ba)

#gets correlation coefficient of EITC and Female_Head
corl_FH <- cor(EITC_Proj$EITC, EITC_Proj$Female_Head)
print(corl_FH)

#gets correlation coefficient of EITC and Poverty_LFPR
corl_PL <- cor(EITC_Proj$EITC, EITC_Proj$Poverty_LFPR)
print(corl_PL)

#gets correlation coefficiet of EITC and SNAP_Part
corl_SNAP <- cor(EITC_Proj$EITC, EITC_Proj$SNAP_Part)
print(corl_SNAP)
```
According the results above, the Single_Women, HS_25, Bachelors_25, Female_Head, Poverty_LFPR, and SNAP_Part variables were statistically significant using a 5% significance level.

The Single_Women variable has a weak positive correlation with the amount of money the government spends on the EITC. Single women take advantage of the EITC program becasue many of them are divorced with children. Furthermore, women tend to take lower paying jobs compared to men, so using the EITC will help their economic situation. The reason the correlation is weak could be because there's a large variety of single women, such as educated women, young women aged 15 to 22, and older widowed women.

The HS_25 variable has a somewhat strong negative correlation with the amount of money the government spends on the EITC. The results indicate that as the number of people with a high school education or higher increases, the amount of government spending on the EITC decreases. The correlation is somewhat strong is possibly due to the variety of socioeconomic classes, marital statuses, family size, etc. The correlation coefficient possibly due to the fact that the more educated individuals are, the higher their salary will be. Thus, a higher salary may disqualify an individual from enrolling in the EITC program.

The Bachelors_25 variable has a very weak positive correlation with the amount of money the government spends on the EITC. The reason why the correlation is weak is probably becasue there's a lot of variety with those who have a bachelor's degree. Those that have a bachelor's degree may have a large family, live in an expensive region such as California, a single parent, etc. The positivity of the correlation coeffiecient may be due to the fact that more educated people may want to take advantage of the EITC because of they tend to be more knowledgeable than lesser educated people.

The Female_Head variable has a somewhat weak positive correlation with the amount of money the government spends on the EITC program. We can infer that the positive correlation probably comes from the fact that single mothers have a difficult time raising children, while holding a job. The tax credit would help single mothers afford the necessities of life. The somewhat weakness may come from the variety of single mothers, such as age, knowledge of the program, disqualified due to being under other governemnt programs, etc.

The Poverty_LFPR has a very weak positive correlation with the amount of money the government spends on the EITC program. We can infer that the strength probably comes from the variety of poor working people. Some poor people may not know the program exists, others may be collecting other benefits from the government. The positive quality may stem from poor people wanting to take advantage of the EITC as more poor people become aware of it.

The SNAP_Part has a very strong positive correlation witht the amount of money the government spends on the EITC program. The strength probably comes from the fact that many SNAP participants are struggling financially, and are more inclinded to enroll in the EITC program to help alleviate their struggles. The positive correlation probably stems from the fact that increasing the number of poor people in a state under multiple government programs, the more government spending is needed for those people.

The code chunk below is to read in the ggplot2 package.
```{r eval = TRUE, echo = TRUE}
#installs ggplot2 package
library(ggplot2)
```

Below is a linear regression and scatter plot of single women and females who are the head of their households.
```{r eval = TRUE, echo = TRUE}
#makes linear regression for Single_Women
Wo_lin <- lm(EITC ~ Single_Women, data = EITC_Proj)

#makes scatter plot for Single_Women
Wo_scat <- ggplot(EITC_Proj, aes(x = Single_Women, y = EITC)) +
  geom_point() + 
  geom_smooth(method = lm)
Wo_scat + 
  labs(title = "Earned Income Tax Credit Given to Single Women",
       x = "Proportion of Single Women",
       y = "Earned Income Tax Credit (in dollars)") +

#adds slope and intercept text on Single_Women scatter plot
annotate('text', x = .4, y = 6*10^9, label = 'Slope: 9.859e+09  Intercept: -3.532e+09')

#makes linear regression for Female_Head
FH_lin <- lm(EITC ~ Female_Head, data = EITC_Proj)

#makes scatter plot with fitted line for Female_Head
FH_scat <- ggplot(EITC_Proj, aes(x = Female_Head, y = EITC)) +
  geom_point() + 
  geom_smooth(method = lm)
FH_scat + 
  labs(title = 'Earned Income Tax credit Given to Females as the Head of Their Households',
       x = 'Female Heads of their Households',
       y = 'Earned Income Tax Credit (in dollars)') +

#adds slope and intercept text on Female_Head scatter plot
annotate('text', x = .16, y = 3.9*10^9, label = 'Slope: 2.154e+10  Intercept: -2.383e+09')

```
As shown above, both scatter plots for female heads of their households and single women have a positive, but weak correlation with the EITC. The points on each scatter plot have a lot of variability and outliers. The outliers are states that have the federal government spending the most money on the EITC program.

Below are linear regression and scatter plot representations of a high school education aged 25 and up.
```{r eval = TRUE, echo = TRUE}
#makes linear regression for HS_25
HS_lin <- lm(EITC ~ HS_25, data = EITC_Proj)

#makes linear scatter plot for HS_25
HS_scat <- ggplot(EITC_Proj, aes(x = HS_25, y = EITC)) +
  geom_point() +
  geom_smooth(method = lm) +
  labs(title = 'EITC Given to Individuals with a High School Education or Higher',
       x = 'Population with a High School Education and Up (age 25 or older)',
       y = 'Earned Income Tax Credit (in dollars)') +

#adds slope and intercept text on HS_25 scatter plot
  annotate('text', x = .935, y = 5.5*10^9, label = 'Slope: -4.140e+10  Intercept: 3.919e+10')

#displays HS_25 scatter plot
print(HS_scat)
```

As shown above, the scatter plot for education is negatively correlated with the EITC variable. The scatter plot captures the scope of the EITC variable because of how it can be interpreted. Simply put, as the proportion of individuals with a high school education or higher increases, the money spent on EITC is shown to decrease, and the reverse is more revealing. The more people with less than a high school education or higher, the more money the government seems to spend on the EITC program. 

Below is a linear regression and scatter plot representation of the SNAP_part variable.
```{r eval = TRUE, echo = TRUE}
#makes linear regression of the SNAP_Part variable
SNAP_lin <- lm(EITC ~ SNAP_Part, data = EITC_Proj)

#makes scatter plot with fitted line for SNAP_Part variable
SNAP_scat <- ggplot(EITC_Proj, aes(x = SNAP_Part, y = EITC)) +
geom_point() +
  geom_smooth(method = lm)
SNAP_scat +
  labs(title = 'EITC Given to Supplemental Nutrition Assistance Program Participants',
       x = 'Supplemental Nutrition Assistance Program Participants',
       y = 'Earned Income Tax Credit (in dollars)') +

#adds slope and intercept text on SNAP_Part scatter plot
  annotate('text', x = .75*10^6, y = 7.5*10^9, label = 'Slope: 3,129  Intercept: 86,156,281')
```
According to the scatter plot above, there is a positive correltion with the SNAP_Part variable. This does not mean that increasing SNAP participants will cause an increase in spending on EITC. However, the scatter plot gives a clue on what types of people to target for EITC spending.

Below is a histogram of participants of SNAP
```{r eval = TRUE, echo = TRUE}
#makes density plot of SNAP_Part
SNAP_hist <- ggplot(EITC_Proj, aes(x = SNAP_Part)) +
geom_histogram(binwidth = 600000)
SNAP_hist +
  labs(title = 'Histogram of the Supplemental Nutrition Assistance Program Participants',
       x = 'Supplemental Nutrition Assistance Program Assistance')
```

The histogram describes a lot about the data. For instance, the histogram is skewed to the right, which may indicate that in the U.S, there is less dependence on government programs, which shows that its economy is successful and thriving. Since there are only a few states that have a high number of SNAP participation, that may indicate that those states may have either a higher population and or poor living conditions. For example, California has a higher population, and also has the highest SNAP participants.

Below is a bar plot of the amount of participants of SNAP and EITC expenditure by state
```{r eval = TRUE, echo = TRUE}
#reorders States and SNAP_Part
EITC_Proj$State <- factor(EITC_Proj$State, levels = EITC_Proj$State[order(EITC_Proj$SNAP_Part)])

#makes bar plot of SNAP_Part
SNAP_bar <- ggplot(EITC_Proj, aes(x = SNAP_Part, y = State))
SNAP_bar + geom_col() +
  labs(title = "Bar Plot of SNAP Participants by State",
       x = "Supplemental Nutrition Assistance Program Participants", y = "State")

#reorders States and EITC
EITC_Proj$State <- factor(EITC_Proj$State, levels = EITC_Proj$State[order(EITC_Proj$EITC)])

#makes bar plot of EITC
EITC_bar <- ggplot(EITC_Proj, aes(x = EITC, y = State))
EITC_bar + geom_col() +
  labs(title = "Bar Plot of EITC Spending on Each State",
       x = "Earned Income Tax Credit (in dollars)", y = "State")
```
According to the bar plot above, the states with the largest amount of SNAP participants are the ones from the most populated states. Furthermore, the states with the most money spent on EITC are the states with the largest populations. Therefore, the distribution of federal money based on the population of a state.

# Works Cited

Philips, Katherin Ross.(2001). Who knows about the earned income tax credit? *The Urban Institute*. https://webarchive.urban.org/publications/310035.html. Accessed Dec. 26, 2023.

Linos, Elizabeth, Allen Prohofsky, Aparna Ramesh, Jesse Rothstein, and Matthew Unrath. (2022) Can Nudges Increase Take-Up of The EITC? Evidence from Multiple Field Experiments. *American Economic Journal:Economic Policy*, 14(4), 432-452. https://doi.org/10.1257/pol.20200603. Accessed Dec. 26, 2023.

Meyer, Bruce D. and Dan T. Rosenbaum. (1999). Welfare, the earned income tax credit, and the labor supply of single mothers. *National Bureau of Economic Research Working Papers*. http://www.nber.org/papers/w7363. Accessed Dec. 26, 2023.

Sage Data. Snap participation – households from the supplemental nutrition assistance program (snap) - monthly database. https://data-sagepub-com.prox.lib.ncsu.edu/dataset?view=AA8BXQAAgABWAQAAAAAAAAAA3_zCT30EyCC3PXJIzZmTt42zIR65nwduP24v7yGJrv_UjkyUnxsR%24w6gMtpE6FS8zWGE_7tB8G9fMnBIcFBGsEjTEh4y_iE8ntJa9LjlLr8lr7B6ZeMn5t3nVRcXAvGUxpSEsMSerO_OzrBcFaeXkvDdFKKBgEct4vZv385TwxDD0leIdqBHhl7BjYgy9zfHZNuffiNxGX305FKCGH2WTe89qxy3O9or0PuYRGVwCA. Accessed Dec. 29, 2023

Sage Data. State and local personal income and employment by major component (non-industry account) from the regional personal income and employment database. (2022). https://data-sagepub-com.prox.lib.ncsu.edu/dataset?view=AA8BXQAAgADJAQAAAAAAAAAA3_zCT30EyCC3PXJIzQwrt429JMW543aSuXEu%24SBt03ghi6FJprzsB8xOXLfPtfwovxyp9yzT8Vjtc85fWBt9wLdJytQhhf3h0%24az0wf%24rZRlo%24Pc0FQJhcX1uIvWcbSZVZi8Kc305wCetn9WplJK77GucvgLJ7vQc%24er6ylF4gMLX4z7U4qH_GL7EPtvvtJFXso7oaytvSN9tuWf4aeH5spOkv3Jeqw7ChYVWNxZY5UmXdWpOCaKM_BoM19ElBZA_ZIcJdxT1Rcg6YA.Dec. 29, 2023.

United States Census Bureau. (2022). Educational Attainment. https://data.census.gov/table?q=education&g=040XX00US01,02,04,05,06,08,09,10,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,44,45,46,47,48,49,50,51,53,54,55,56. Accessed Dec. 29, 2023.

United States Census Bureau. (2022). Employment status. https://data.census.gov/table?q=employment%20status&g=040XX00US01,02,04,05,06,08,09,10,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,44,45,46,47,48,49,50,51,53,54,55,56. Dec. 29, 2023.

United States Census Bureau. (2022). Households and families. https://data.census.gov/table?q=Families%20and%20Household%20Characteristics&g=010XX00US_040XX00US01,02,04,05,06,08,09,10,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,44,45,46,47,48,49,50,51,53,54,55,56. Accessed Dec. 29, 2023.

United States Census Bureau. (2022). Income in the past 12 months (in 2022 inflation-adjusted dollars). https://data.census.gov/table?q=income&g=010XX00US_040XX00US01,02,04,05,06,08,09,10,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,44,45,46,47,48,49,50,51,53,54,55,56. Accessed Dec. 29, 2023.

United States Census Bureau. (2022). Marital status. https://data.census.gov/table?q=marital%20status&g=040XX00US01,02,04,05,06,08,09,10,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,44,45,46,47,48,49,50,51,53,54,55,56. Accessed Dec. 29, 2023.