---
title: "HW2MD"
author: "Jiordani Etienne"
date: "3/3/23"
output:
  html_document:
    df_print: paged
---

#Task 1

1.  What is the main object that the tidyverse functions act on? What Base R data structure is it most similar to?

2.  Why do we like to you chaining when possible as opposed to the standard way of calling R functions?

Chaining can help the code be more readable

3.  What is meant by 'long-form' data? Which package from the tidyverse allows us to transform data from wide to long format (and vice-versa)?

Long-form data is when a dataset has a few variables arranged shortly, and observations are arranged to have more vertical distance. Use pivot_longer(). To switch back to wide for, use pivot_wider().

# Task 2

## 1. I first write the code to read in the excel file. Next, I write in r markdown the code chunk:

read in a jretienn_beans

```{r echo = TRUE}
#read in jretienn_beans
library(tidyverse)
library(readxl)
beans <- read_excel('jretienn_beans.xlsx')

#identify beans
str(beans)
```

The data has been read into the object beans. This object is a dataframe that has `r length(beans)` variables and `r length(beans$Area)` observations.

## 2. Use tidyverse and chaining to change the dataset

```{r echo = TRUE, eval = TRUE}
#exclude Convexarea and EquivDiameter
beans2 <- beans %>% select(-'ConvexArea',-'EquivDiameter') %>%

#rename ShapeFactor1, ShapeFactor2, ShapeFactor3 and ShapeFactor4 
rename('SF1' = ShapeFactor1, 'SF2' = ShapeFactor2, 'SF3' = ShapeFactor3, 'SF4' = ShapeFactor4) %>%

#filter class
filter(Class %in% c('HOROZ', 'DERMASON', 'SEKER', 'BOMBAY', 'SIRA')) %>%

#create a variable with the average
mutate(avg = (SF1 + SF2 + SF3 + SF4)/4)%>%

#create categorical variables of Roundness 
mutate(size = if_else(Roundness >=  0.909371, 'large', 
                        if_else(Roundness >= 0.8721575, 'medium', 'small')))%>%

#reorder rows
arrange(desc(Roundness))
```

# Task 3

## 3 I subset the data

```{r echo = TRUE}
#subset data
beans3 <- beans2 %>% select('Class', 'size', 'Roundness', 'AspectRatio', 'Solidity')%>%

#filter for only medium and small variables show
filter(size %in% c('medium', 'small'))
```

The first value of the Class variable is `r beans3$Class[1]` and the first value of Roundness is `r beans3$Roundness[1]`.
