---
title: "HW3"
author: "Jiordani Etienne"
date: '`30 March 2023'
output:
  html_document: default
  pdf_document: default
---

#Conceptual Questions

##1. What are the common plots and numeric summaries we use to describe a numeric variable’s distribution?

Plots and numerical summaries that describe a numerica variable's distribution are contingency tables, mean, median, standard deviation, IQR, quantiles, and percentiles. We also use bar plots, histograms, box plots, and scatter plots.

##2. What is meant by the term ‘two-way contingency table’?

A two-way contingency table is a table of a combination of two categorical variables that describes the pattern and frequences the observations.

#Programming Questions

##1. Replicate data and modify the code
```{r echo = TRUE}
#read in jretienn_beans
library(tidyverse)
library(readxl)

beans <- read_excel('jretienn_beans.xlsx')

#exclude Convexarea and EquivDiameter
beans2 <- beans %>% select(-'ConvexArea',-'EquivDiameter') %>%

#rename ShapeFactor1, ShapeFactor2, ShapeFactor3 and ShapeFactor4 
rename('SF1' = ShapeFactor1, 'SF2' = ShapeFactor2, 'SF3' = ShapeFactor3, 'SF4' = ShapeFactor4) %>%

#filter class
filter(Class %in% c('HOROZ', 'DERMASON', 'SEKER', 'BOMBAY', 'SIRA')) %>%

#create a variable with the average
mutate(avg = (SF1 + SF2 + SF3 + SF4)/4)%>%

#create categorical variables of Roundness 
mutate(size = if_else(Roundness >=  0.909371, 'large', 
                        if_else(Roundness >= 0.8721575, 'medium', 'small')))%>%

#reorder rows
arrange(desc(Roundness))

#subset data
beans3 <- beans2 %>% select('Class', 'size', 'Roundness', 'AspectRatio', 'Solidity')%>%

#filter for only medium and small variables show
filter(size %in% c('medium', 'small'))
```

##2. Create a contengincy table of Class and size variable
```{r echo = TRUE, eval = TRUE}
#create a new dataset and filter it
beans4 <- beans2 %>% select('Class', 'size', 'SF1', 'Area', 'Perimeter', 'MajorAxisLength', 'MinorAxisLength', 'AspectRatio', 'Eccentricity') %>%

#include all observations that have SF1 less than .0081
filter(SF1 < .0081)

#Create the contingency table
table(beans4$Class, beans4$size)
```
The number 12 on the top left means that there are 12 seeds that are large, BOMBAY, and a shape factor 1 of under .0081

##3. Find the Skew, miniumum, 5th percentile for Area and Perimeter variables
```{r echo = TRUE, eval = TRUE}
#make a summary of beans4

#bring in skewness function
library(e1071)
beans4%>%
group_by(Class)%>%

#summarize Area and Perimeter by making formulas
summarize(avga = mean(Area),
          skewa = skewness(Area),
          mina = min(Area),
          quaa = quantile(Area, probs = .05),
          avgp = mean(Perimeter),
          skep = skewness(Perimeter),
          minp = min(Perimeter),
          quap = quantile(Perimeter, probs = .05))
```
For the BOMBAY class, avga is the average area of all BOMBAY. skewa is the skewness of all BOMBAY seeds. mina is the minimum area out of all of the BOMBAY observations. quaa is the bottom 5% area of all of the BOMBAY. avgp is the average perimeter of all BOMBAY observations. skep is the skewness of all of the BOMBAY.minp is the minimum perimeter from the BOMBAY observations. quap is the bottom 5% of the perimeter of the BOMBAY observations.

##4. Create correlation between variables
```{r echo = TRUE, eval = TRUE}
#subset data
beans5 <- beans4 %>% select('Area', 'Perimeter', 'MajorAxisLength', 'MinorAxisLength', 'AspectRatio')

#create a correlation
cor(beans5)
```

##5. Make a side-by-side barplot for Class and size variables
```{r echo = TRUE, eval = TRUE}
#create a side-by-side barplot of Class and size variables and add a label
ggplot(data = beans4, aes(x = Class, fill = size)) + geom_bar(position = 'dodge') + labs(title = 'Seed size for every seed type')
```

##6. Make a histogram of the Eccentricity variable, make a density plot, and add a title
```{r echo = TRUE, eval = TRUE}
#make a histogram of Eccentricity
ggplot(data = beans4, aes(x = Eccentricity)) + geom_histogram(aes(y = ..density..)) + 
  
#make a density plot
  geom_density() +

#make a title
  labs(title = 'Histogram of Eccentricity')
```

##7. Create a histogram at every level of Class
```{r echo = TRUE, eval = TRUE}
#make a histogram of Eccentricity
ggplot(data = beans4, aes(x = Eccentricity)) + geom_histogram(aes(y = ..density..)) + 
  
#make it at each level of Class
facet_wrap(~Class) +
  
#make a density plot
geom_density() +

#make a title
  labs(title = 'Histogram of Eccentricity')
```

